Mathematical Aspects of the Capital Market

### 2017-08-07 phd/Chapterintroduction.tex
progress on the importance sampling paper. Added to the introduction section of the thesis.

\chapter{Introduction}\label{ch:intr}
The phenomena of heavy-tailedness is widely observed in many
disciplins of science, for example, phase transition of matter and
black body radiation as studied in physics, neuronal avalanches in
biology, claim sizes of insurance mathematics and stock returns in
finance. The last application is indeed the focus of this thesis. To
discuss the phenomena in precise terms, we introduce the concept of
regular variation.

@@

@@

\section{Stochastic Recurrence Equation}
One of the most important dynamical mechanisms that lead to regularly
varying r.v. is stochastic recursion of the following form:
\[
X_t = A_t X_{t-1} + B_t
\]
where $X_t$ is a $d$-dimensional random vector, $A_t$ is a $d\times d$
random matrix and $B_t$ is a $d$-dimensional vector, random or
deterministic. The sequence $\{A_i\}_{i=1,2,\dots}$ and
$\{B_i\}_{i=1,2,\dots}$  are iid and independent of each other.

@@

Recently, Collamore and Mentemeier \cite{collamore:mentemeier:2016}
extended Kesten's result and gave an explicit expression for $\mu_\xi$:
\begin{equation}
  \label{eq:CollamoreMentemeierIntro}
  \lim_{u \to \infty} u^{\xi} \E \left[
    f(u^{-1} V)
    \right]
  =
  {C \over \lambda'(\xi)}  
  \int_{\sphere^{d-1}_+ \times R} e^{-\xi s} f(e^s x) \ell_\xi(dx) ds
\end{equation}
where $C$ is a constant (cf. (2.15) of Collamore and Mentemeier
\cite{collamore:mentemeier:2016}), $f(\cdot)$ is any bounded
continuous function on $\reals^{d}_+ \setminus \{0\}$  and
$\ell_\xi$ is a probability measure on $\sphere_+^{d-1}$. Its definition
is also found in \eqref{eq:eigenmeasure}.

From \eqref{eq:CollamoreMentemeierIntro} a representation for $\mu_\xi
(\cdot)$ immediately follows  
\[
\mu_\xi (\cdot) = {C \over \lambda'(\xi)} \mathcal L_\xi(\cdot)
\]
Here $\mathcal L_\xi$ is a non-null Radon measure  on
$\reals^d_+ \setminus \{0\}$ that satisfies, for all
bounded continuous function $f(\cdot)$ on
$\reals_+^{d} \setminus \{0\}$:
\[
\int_{\reals_+^d\setminus \{0\}} f(x) \mathcal L_\xi(dx)
=
\int_{\sphere^{d-1}_+ \times R} e^{-\xi s} f(e^s x) \ell_\xi(dx) ds
\]

>>>>

\chapter{Introduction}\label{ch:intr}
The phenomena of heavy-tailedness is widely observed in many
disciplines of science, for example, phase transition of matter and
black body radiation as studied in physics, neuronal avalanches in
biology, claim sizes of insurance mathematics and stock returns in
finance. The last application is indeed the focus of this thesis. To
discuss the phenomena in precise terms, we introduce the concept of
regular variation.

@@

Figure \ref{fig:thjyuj} shows the Hill estimates of the tail indices
of daily stock return series from 3 sectors of the
{\em Standard \& Poor 500} index \footnote{An American stock index
  comprising 500 or so companies.}. The 2.5\% and 97.5\% quantiles of
the asymptotic normal distribution of the estimates are also
given. One can see the confidence bands are fairly large compared with
the estimated values. This certainly raises the question of how
similar they really are and if/how their variations can be explained by
economical arguments. We explore this topic in chapter
\ref{ch:TailParameters}.

\begin{figure}[htb!]
  \begin{minipage}{1.0\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
    {Energy_lower.pdf}
  \end{minipage}
  \begin{minipage}{1.0\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
    {Consumer_Staples_lower.pdf}
  \end{minipage}
  \begin{minipage}{1.0\linewidth}
    \includegraphics[width=\textwidth, trim={0, 0.8cm, 0, 2cm}, clip]
    {Information_Technology_lower.pdf}
  \end{minipage}
  \caption{\small Hill estimates $\hat \alpha_{50}$ of the lower
    tail-indices $\alpha$ of daily return series in sectors of the S\&P 500
    index. The data span from 1 January 2010 to 31 December 2014 and
    comprise $n=1304$ observations.
    The graphs from top to bottom correspond to the ``Energy'',
    ``Consumer Staples'' and ``Information Technology'' sectors.
    Each circle corresponds to a Hill estimate $\hat\alpha_{50}$; the gray
    triangles above and below it mark the 97.5\% and 2.5\% quantiles
    of its approximate normal distribution; see \eqref{eq:2} and the
    discussion following it for an interpretation.
    The lower and upper red lines mark the medians of the 2.5\% 
    and 97.5\% quantiles, respectively, evaluated from all stocks in
    the sector.
    The data are taken from {\it Yahoo Finance}; the labels on
    the horizontal axes are Yahoo symbols of the stocks. 
  }\label{fig:thjyuj}
\end{figure}

Random variables with regularly varying tails have some very nice
features: if $X_1$ and $X_2$ are both regularly varying with indices
$\alpha_1$ and $\alpha_2$, respectively, then $a_1 X_1 + a_2 X_2$ is
regularly varying with index $\min\{\alpha_1, \alpha_2\}$. Moreover
if $X_1, X_2$ are iid,
$\P(a_1 X_1 + a_2 X_2 > u) \sim \P(a_1 X_1 > u) + \P(a_2 X_2 > u)$.

Now consider $N$ return series $X_{i,t}$, $i=1,2,\dots, p$. Suppose
each of these series is a linear combination of $K$ factors $Y_{i,t}$,
$i=1,2,\dots,K$, the $i$-th of which is regularly varying with index
$\alpha_i$. Then by the summation property, each and every $X_i$ is
regularly varying with index $\min_{1 \leq i \leq K} \alpha_i$.
Since in practice a factor $Y_{i,t}$ is found as a linear combination
of $X_{i,t}$, the observed time series, following an eigenvector of
the sample covariance matrix of $X_{i,t}$, it is important to
understand the eigensystem of this matrix. This topic indeed
constitutes chapters \ref{ch:bernoulli} and \ref{ch:extremes} of this
thesis.

When a product of random variables, say $X_1 X_2$, involves a random
variable with regularly varying tails, a useful result is that of
Breiman: assume $X_1$ is regularly varying with index $\alpha$ and
there exists $\epsilon > 0$ such that
$\E |X_2|^{\alpha + \epsilon} < \infty$.
Then $\P(X_1 X_2 > x) \sim \E|X_2|^{\alpha} \P(X_1 > x)$.
More generally, if $X_1, X_2$ are regularly varying with the same tail
index $\alpha$ or if $\P(X_2 > x) = o(\P(X_1 > x))$, then $X_1 X_2$ is
regularly varying with index $\alpha$.

For an extensive summary of the regular variation properties of
functions of regularly varying random variables, see Mikosch and
Jessen \cite{JessenMikosch2006}.

@@

\section{Stochastic Recurrence Equation}
One of the most important dynamical mechanisms that lead to regularly
varying r.v. is stochastic recursion of the following form:
\begin{equation}
  \label{eq:rhjyu}
  X_t = A_t X_{t-1} + B_t
\end{equation}
where $X_t$ is a $d$-dimensional random vector, $A_t$ is a $d\times d$
random matrix and $B_t$ is a $d$-dimensional vector, random or
deterministic. The sequence $\{A_i\}_{i=1,2,\dots}$ and
$\{B_i\}_{i=1,2,\dots}$  are iid and independent of each other.

@@

%% Recently, Collamore and Mentemeier \cite{collamore:mentemeier:2016}
%% extended Kesten's result and gave an explicit expression for $\mu_\xi$:
%% \begin{equation}
%%   \label{eq:CollamoreMentemeierIntro}
%%   \lim_{u \to \infty} u^{\xi} \E \left[
%%     f(u^{-1} V)
%%     \right]
%%   =
%%   {C \over \lambda'(\xi)}  
%%   \int_{\sphere^{d-1}_+ \times R} e^{-\xi s} f(e^s x) \ell_\xi(dx) ds
%% \end{equation}
%% where $C$ is a constant (cf. (2.15) of Collamore and Mentemeier
%% \cite{collamore:mentemeier:2016}), $f(\cdot)$ is any bounded
%% continuous function on $\reals^{d}_+ \setminus \{0\}$  and
%% $\ell_\xi$ is a probability measure on $\sphere_+^{d-1}$. Its definition
%% is also found in \eqref{eq:eigenmeasure}.

%% From \eqref{eq:CollamoreMentemeierIntro} a representation for $\mu_\xi
%% (\cdot)$ immediately follows  
%% \[
%% \mu_\xi (\cdot) = {C \over \lambda'(\xi)} \mathcal L_\xi(\cdot)
%% \]
%% Here $\mathcal L_\xi$ is a non-null Radon measure  on
%% $\reals^d_+ \setminus \{0\}$ that satisfies, for all
%% bounded continuous function $f(\cdot)$ on
%% $\reals_+^{d} \setminus \{0\}$:
%% \[
%% \int_{\reals_+^d\setminus \{0\}} f(x) \mathcal L_\xi(dx)
%% =
%% \int_{\sphere^{d-1}_+ \times R} e^{-\xi s} f(e^s x) \ell_\xi(dx) ds
%% \]

\section{GARCH models}
Introduced by Bollerslev in 1986, GARCH models have been hugely
popular in modelling volatility of financial time series and have
inspired numerous variants. Its original form is the following the
recurrence equation:
\[
\sigma_t^2 = \omega + \sum_{i=1}^p \alpha_i R_{t-i}^2 +
\sum_{j=1}^q \beta_j \sigma_{t-j}^2
\]
where $\R_t$ is a return series, e.g. stock returns, foreign exchange
rates, interest rates, etc; $\sigma_t^2$ is the variance of the
distribution of $R_t$ conditional on $\{(R_i,
\sigma_i^2)\}_{i=0}^{t-1}$. $\omega, \{\alpha_i\}_{i=1}^p,
\{\beta_i\}_{i=1}^q$ are constant parameters of the model. Clearly,
the GARCH($p,q$) recurrence equation is of the form of
\eqref{eq:rhjyu}. So, with appropriate conditions, including
$\sum_{i=1}^p \alpha_i + \sum_{j=1}^q \beta_j < 1$, $\sigma_t^2$ is
shown to be a positive Harris Markov chain (cf. Bollerslev
\cite{bollerslev:1986} and Buraczewski et al
\cite{buraczewski:damek:mikosch:2016}), whose stationary distribution
has regularly varying tails. The tail index, call it $\xi$, is given by
\[
\lim_{n \to \infty} {1 \over n}\log\E\|A_n \cdots A_1\|^\xi = 0
\]
where $A_1, A_2, \dots, A_n$ are iid matrices whose entries are
functions of $\{\alpha_i\}_{i=1}^p$ and $\{\beta_i\}_{i=1}^q$:
\[
A_i =
\begin{pmatrix}
  \alpha_1 Z_{t-1}^2 + \beta_1 & \beta_2 & \cdots &
  \beta_{q-1} & \beta_q & \alpha_2 & \alpha_3 &
  \cdots & \alpha_{p-1} & \alpha_p\\
  1 & 0 & \cdots & 
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots & 
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  1 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  Z_{t-1}^2 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  0 & 0 & 1 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots &
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\    
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 1 & 0 \\    
\end{pmatrix}
\begin{pmatrix}
  \sigma_{t-1}^2 \\
  \sigma_{t-2}^2 \\
  \vdots \\
  \sigma_{t-q+1}^2 \\
  \sigma_{t-q}^2 \\
  R_{t-2}^2 \\
  R_{t-3}^2 \\
  \vdots \\
  R_{t-p+1}^2 \\
  R_{t-p}^2
\end{pmatrix}
\]
